Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*

### 2020-06-06 paper.Rmd

---
title: "Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
author: Andreas Beger and Richard K. Morgan and Michael D. Ward
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  pdf_document:
    keep_tex: true
references:
- id: blair2020forecasting
  title: "Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning"
  author:
  - family: Blair
    given: Robert A.
  - family: Sambanis
    given: Nicholas
  container-title: Journal of Conflict Resolution
  type: article-journal
  issued:
    year: 2020
bibliography: 
  - "`r user <- Sys.info()[['user']]; if (user=='andybega')  '/Users/andybega/Work/whistle/master.bib' else if (user=='mdw') '/Users/mdw/Git/whistle/master.bib' else 'references.bib'`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE)

@@

To examine the first question, B&S compare the test set of the escalation model to four alternative models. The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country. The models are:
- Escalation: a set of ten indicators, putatively drawn from a theoretical escalation model.
- Quad: ICEWS quad counts, i.e. material conflict, material cooperation, verbal conflict, verbal cooperation.
- Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for interactions between the government one one side and opposition or rebel actors on the other. These are directed, thus making for four total covariates.
- CAMEO: counts for all CAMEO event codes totalling $1,159$ covariates, which are mostly zero for any country in any month. 
- Average: unweighted average of the predictions from the four models briefly described above.

@@

While replicating and analyzing B&S's results, we found several issues worthy of further discussion and investigation.
These are a) the use of smoothed ROC curves to draw conclusions about which model is best, b) an incorrect implementation of the weighted by ICEWS data model, c)  an inconsistent test set $ for the models examined, d) incorrect scoring of the 2016 forecasts, e) an inappropriate use of random forrest implementation, with very unusual hyperparameters.  Our belief is that these research decisions lead B&S to draw conclusion that are incorrect. The escalation model is not the best and it actually performs worse that the atheoretical, garbage can model with over $1000$ variables.  We turn to discussing these five issues below.  We hold a complete analysis that corrects all these issues until later, as there are many possible permuations of a seriatim unfolding.

>>>>

---
title: "Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*"
author: Andreas Beger, Richard K. Morgan, and Michael D. Ward
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  pdf_document:
    keep_tex: true
references:
- id: blair2020forecasting
  title: "Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning"
  author:
  - family: Blair
    given: Robert A.
  - family: Sambanis
    given: Nicholas
  container-title: Journal of Conflict Resolution
  type: article-journal
  issued:
    year: 2020
bibliography: 
  - "`r user <- Sys.info()[['user']]; if (user=='andybega')  '/Users/andybega/Work/whistle/master.bib' else if (user=='mdw') '/Users/mdw/Git/whistle/master.bib' else 'references.bib'`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE)
# options(tinytex.verbose = TRUE)

@@

To examine the first question, B&S compare the test set of the escalation model to four alternative models. The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country. The models are:
- Escalation: a set of ten indicators, putatively drawn from a theoretical escalation model.
- Quad: ICEWS quad counts, i.e. material conflict, material cooperation, verbal conflict, verbal cooperation.
- Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for interactions between the government one one side and opposition or rebel actors on the other. These are directed, thus making for four total covariates.
- CAMEO: counts for all CAMEO event codes totaling $1,159$ covariates, which are mostly zero for any country in any month. 
- Average: unweighted average of the predictions from the four models briefly described above.

@@

While replicating and analyzing B&S's results, we found several issues worthy of further discussion and investigation.
These are a) the use of smoothed ROC curves to draw conclusions about which model is best, b) an incorrect implementation of the weighted by ICEWS data model, c)  an inconsistent test set for the models examined, d) incorrect scoring of the 2016 forecasts, e) an inappropriate use of random forrest implementation, with very unusual hyperparameters.  Our belief is that these research decisions lead B&S to draw conclusion that are incorrect. The escalation model is not the best and it actually performs worse that the atheoretical, garbage can model with over $1000$ variables.  We turn to discussing these five issues below.  We hold a complete analysis that corrects all these issues until later, as there are many possible permuations of a seriatim unfolding.
